### 2018-03-08 chapter3/ms.tex
Chapter 3: Second Pass at References. Bulk Formatting

\section{Introduction}
\label{sec:ch3_intro}
The detection of gravitational wave (GW) events during the first and second observing runs of the Advanced Laser Interferometer Gravitational-Wave Observatory (aLIGO) ushered in the era of GW astronomy. The first four announced detections (GW150914, GW151226, GW170104, and GW170814) were due to binary black hole (BBH) mergers, with component masses of $\approx8-36$ M$_\odot$ \citep{LIGOGW150914,LIGOGW151226,LIGOGW170104,LIGOGW170814}.

More recently LIGO announced the first detection of gravitational waves from the inspiral and merger of a binary neutron star \citep[GW170817][]{LIGOGW170817}. This remarkable event was accompanied 1.7s later by a coincident short gamma-ray burst (SGRB) detected with the {\it Fermi} and {\it INTEGRAL} satellites \citep{LIGOGW170817grb,GW170817Fermi,Savchenko+17}. An intense follow-up effort led to the discovery of an associated optical counterpart just 11 hours post-merger \citep{LIGOMMAPaper,gw170817lco,Coulter+17,GW170817DECam,Valenti+17}.

Initial modeling of this optical emission revealed that it was consistent with the behavior expected for a kilonova \citep{cowp17, kilpatrick17, Tanaka+18, villar17}, an isotropic thermal optical/NIR transient powered by the radioactive decay of $r$-process nuclei synthesized in the merger ejecta \citep[see e.g.,][]{Metzger+10,Kasen+13,BarnesKasen13,Metzger2017}. Subsequent studies into the EM counterpart revealed the nature of the radio and X-ray emission \citep{Alexander+17,davanzo18,Margutti+17,Margutti+18,Mooley+18,Troja+17} and confirmed the notion of compact object mergers as the progenitors of SGRBs \citep{Fong+13, Berger2014, Fong+15, Fong+17}.

While the case of GW170817 represents a clear success and the dawn of joint GW-EM astronomy, there are many unique aspects of this discovery that should be considered. First, the source was unusually close \citep[$D \approx 40$~Mpc,][]{LIGOGW170817} with a clearly associated host galaxy \citep[NGC 4993,][]{Blanchard+17,cantiello18}. Furthermore, the source was unexpectedly bright at discovery \citep[$m_i \approx 17.5$~mag,][]{LIGOMMAPaper,gw170817lco,Coulter+17,GW170817DECam,Valenti+17} making searches with 1-m class telescopes feasible. It is not clear that future detections by aLIGO will be so fortuitous or if the observed features of the kilonova associated with GW170817 will be ubiquitous. Therefore, in the context of future searches for EM counterparts, wide-field instruments on large telescopes will still be necessary to perform efficient and effect follow-up of GW sources.

For wide-field searches conducted in the optical bands, kilonovae like the one associated with GW170817 are of particular interest due to the isotropic nature of the emission and the expectation that this emission will accompany all BNS mergers, and some of the NS-BH mergers in which the NS is disrupted outside the event horizon \citep{MetzgerBerger12}. The short duration ($\lesssim {\rm week}$) and low luminosity ($L_{\rm bol} \sim {\rm a few} \times 10^{41}$~erg s$^{-1}$ at peak) of the kilonova emission make deep and rapid searches imperative. Coupled with the large search regions, this also requires the development of methodologies to robustly identify kilonovae among the vast numbers of potential contaminating sources. For example, optical follow-up of GW151226, covering hundreds of square degrees, has led to the identification of tens of unrelated transient sources despite being shallower than ideal searches for kilonova emission require \citep[e.g.,][]{cenko15,cenko16,GW151226PS1}. Similar wide-field efforts for GW170817 \citep[e.g., DECam][]{GW170817DECam} found 81 transients in a search covering 70 deg$^2$.

@@

The paper is organized as follows. In \cref{sec:ch3_kn_models} we describe the various kilonova models explored in this work. We present the observations and data analysis in \cref{sec:ch3_obs}. In \cref{sec:ch3_search} we discuss the methodologies and results of our search for kilonova contaminants. In \cref{sec:ch3_fakes}, we determine our search efficiency using point sources injected into the images with a wide range of brightness, fading rate, and color. In \cref{sec:ch3_contam_rates} we determine the contamination rate as a function of $i-z$ color and compare our results to those reported from optical follow-up of the current BBH merger events. We discuss the implications of our search in the context of current and future GW detectors and follow-up facilities in \cref{sec:ch3_conc}.

@@

Radioactive decay of the $r$-process elements synthesized during the merger heats the ejecta producing an isotropic, thermal transient \citep{LP98,Rosswog2005, Metzger2010,Tanaka+14}. The combination of low ejecta mass and high ejecta velocity, coupled with the strong optical line blanketing, results in a transient that is faint ($i \approx 23$ and $z \approx 22$ mag at 200 Mpc), red ($i-z \gtrsim 0.5$ mag), and short-lived with a typical duration of $\sim {\rm few}$ days in $z$-band and $\sim{\rm week}$ in $J$-band \citep{BarnesKasen13,Barnes+16,Metzger2017}. In the case of larger opacities \citep[e.g.,][]{Fontes+17}, the transient is expected to peak in the IR (\apx3~$\mu$m), with a duration of \apx10~days \citep{Fontes+17,Wollaeger+17}.

@@

To summarize, kilonova emission with red color $(i-z\gtrsim 0.5)$, a peak brightness of $z\approx22.2$~mag at 200 Mpc, and a duration of $\sim \rm few$ days is expected to be ubiquitous. Blue kilonova emission due to a surviving HMNS is expected to be bluer ($i-z \lesssim 0$~mag), similarly faint ($r \approx 22$~mag) and shorter in duration ($\lesssim 1-2$ days). This picture was confirmed by modeling of the optical emission associated with GW170817. Successfully fitting the data required models comprised of at least two components (e.g., `red' and `blue' emission) with their own distinct ejecta properties \citep{cowp17, kilpatrick17, Tanaka+18, villar17}. The observed behavior and composition of the dynamical ejecta and associated red kilonova emission is robust and ubiquitous, however the nature and observability of the blue kilonova emission depends both on the fraction of cases in which a HMNS survives (unknown) and on geometrical effects ($\lesssim 50\%$).  Finally, emission due to free neutrons will be similar in brightness and color to the blue kilonova emission, but with a timescale of only $\sim \rm few$ hours; the prevalence of this signal is uncertain.  The observations described in this paper address contamination in all of these cases.

@@

We processed the data using {\tt photpipe}, an image processing pipeline used by several previous time domain surveys (see e.g., \citealt{Rest+05,Rest+14}) to perform single-epoch image processing, image subtraction, and candidate identification. The single epoch processing began with the raw images and appropriate calibration files obtained from the NOAO archive\footnote{\url{http://archive.noao.edu/}} and initial image reduction (e.g., bias subtraction, cross-talk corrections, flat-fielding). We performed astrometric calibration relative to the 2MASS $J$-band point-source catalog. Search and template images were deprojected into a tangential plane and pairs of $i$- and $z$- band images from a single epoch were coadded using {\tt SWARP} \citep{Bertin+02}.

@@

\begin{deluxetable}{lcc}
\tabletypesize{\footnotesize}
\tablecolumns{3}
\tablewidth{0pt}
\captionsetup{justification=centering}
\tablecaption{Field Pointing Centers
\label{tab:ch3_pointings}}
\tablehead{
    \colhead{Pointing} &
    \colhead{R.A.} &
    \colhead{Decl.} \\}
\startdata
Antlia 1 & 10:19:58.5  & $-$30:55:24 \\
Antlia 2 & 10:19:44.5  & $-$33:07:24 \\
Antlia 3 & 10:19:30.5  & $-$35:19:24 \\
Antlia 4 & 10:19:10.5  & $-$37:31:24 \\
Antlia 5 & 10:30:03.5  & $-$30:55:24 \\
Antlia 6 & 10:30:03.5  & $-$33:07:24 \\
Antlia 7 & 10:30:03.5  & $-$35:19:24 \\
Antlia 8 & 10:30:03.5  & $-$37:31:24 \\
Antlia 9 & 10:40:08.5  & $-$30:55:24 \\
Antlia 10 & 10:40:22.5  & $-$33:07:24 \\
Antlia 11 & 10:40:38.5  & $-$35:19:24 \\
Antlia 12 & 10:40:57.5  & $-$37:31:24 \\
Antlia 13 & 10:50:12.5  & $-$30:55:24 \\
Antlia 14 & 10:50:41.5  & $-$33:07:24 \\
Antlia 15 & 10:51:13.5  & $-$35:19:24 \\
Antlia 16 & 10:51:50.5  & $-$37:31:24 \\
Antlia 17 & 11:00:17.5  & $-$30:55:24 \\
Antlia 18 & 11:01:00.5  & $-$33:07:24 \\
Antlia 19 & 11:01:48.5  & $-$35:19:24 \\
Antlia 20 & 11:02:44.5  & $-$37:31:24 \\
Antlia 21 & 10:36:36.0  & $-$27:31:04 \\
\enddata
\tablecomments{Central J2000 coordinates for all 21 fields used in this analysis.}
\end{deluxetable}

In addition to the photometry performed by {\tt photpipe}, we also constructed secondary catalogs for all sources identified in the $griz$-band template epochs. This was accomplished using the {\tt Source Extractor} ({\tt SExtractor}) photometry package in single-image mode \citep{BertinArnouts96}. We also performed forced aperture photometry in the template epoch at the position of each candidate identified by {\tt photpipe}. This approach helps to identify the presence of flux in the template images for objects not detected by {\tt SExtractor} which is indicative of image artifacts and defects. We use these additional template catalogs as a useful tool for candidate classification and artifact rejection (see \cref{sec:ch3_source_cuts}).

@@

\section{A Search for Optical Transients}
\label{sec:ch3_search}

@@

We note that at the time these data were taken CCD \#61 (N30) was not functioning. Additionally, during processing of the candidate list we observed severe data quality issues with CCDs \#16 (S17) and \#44 (N13) that produced a number of image artifacts and erroneous detections several orders of magnitude larger than in the other CCDs. Consequently, these three CCDs have been excluded from the analysis. This results in a $\approx 5\%$ loss of sky coverage as discussed in \cref{tab:ch3_search}.

@@

In Groups 3 and 4, the choice of galaxy brightness and matching radius are motivated by observations of the host galaxies of Short GRBs, which have luminosities of $\gtrsim 0.1$ L$^\star$ \citep{FongBerger13,Fong+13,Berger2014}. At the ALV design-sensitivity detection range $(\sim 200$ Mpc) such galaxies will be brighter than 17.5 mag. We then select the matching radius such that the probability of the point source and galaxy being associated by chance, the ``probability of chance coincidence", is $P_{\rm cc}\lesssim 0.1$. This is given by:

@@

\begin{figure}[!t]
\begin{center}
\hspace*{-0.1in}
\scalebox{1.}
{\includegraphics[width=0.4\textwidth]{./figs/chapter3/f2.pdf}}
\caption{{\it Top:} $r-i$ vs. $i-z$ color-color diagram for the template counterparts identified in Groups 3 (orange stars; \cref{sec:ch3_group3}) and 4 (blue circles; \cref{sec:ch3_group4}). The black contours indicate the stellar locus in our images. We find that the majority of sources are consistent with the stellar locus indicating a variable star origin.
{\it Middle:} Same as top, but only showing those 14 sources that do not have a match in the Gaia DR1 catalog. The majority of sources are still consistent with the stellar locus, indicating that they are stellar but simply too faint to appear in the Gaia catalog.
{\it Bottom:} Distribution of $r$-band magnitudes for sources in Groups 3 and 4. The solid lines indicate sources that have a counterpart in the Gaia DR1 catalog, while dashed lines indicate sources that do not have a catalog match.}
\label{fig:ch3_color_color}
\end{center}
\end{figure}

@@

We note that the single source in \Cref{fig:ch3_offset_hist} with an offset of $\approx6.6\arcsec$, was originally identified with an offset of $0.14\arcsec$. However, manual inspection revealed that the transient had not completely faded away in the template epoch causing {\tt SExtractor} deblending to detect a low stellarity point source still present on top of the galaxy light distribution. We correct this offset manually, but leave this source in Group 2 as that is the {\it original} classification.

\begin{figure}[!t]
\begin{center}
\hspace*{-0.1in}
\scalebox{1.}
{\includegraphics[width=0.9\textwidth]{./figs/chapter3/f3.pdf}}
\caption{{\it Left:} $i-z$ color distribution for the 48 sources considered in this analysis. The 11 sources identified as red kilonova contaminants are plotted in orange (\cref{sec:ch3_kn_search}), while the 10 sources identified as blue kilonova contaminants are shown in blue (\cref{sec:ch3_kn_search}). The remaining 27 sources from Group 2 are plotted in grey dashed lines.
{\it Right:} The number of sources recovered by performing an $i-z$ color cut on the sample of 48 sources in Group 2. The orange line gives the number of sources redder than a given $i-z$ color. The blue line indicates the number of sources bluer than a given color cut. The vertical black lines indicate our nominal cuts of $i-z < 0$~mag and $i-z > 0.5$~mag, for blue and red kilonovae, respectively. The sharp increase in the number of sources as the chosen color threshold is relaxed can be clearly seen.}
\label{fig:ch3_color_dist_final}
\end{center}
\end{figure}

\subsubsection{Groups 3 and 4 -- Point Source Counterparts}
\label{sec:ch3_pointsources}
We are also interested in those candidates that have a coincident point source counterpart in the template images. The presence of such a counterpart disqualifies the source as a kilonova contaminant, but this determination relies on the existence of late-time (or pre-existing) templates.  Therefore, it is still meaningful to construct a census of these sources to gain a better understanding of potential contamination in real-time GW follow-up observations, especially if pre-existing template images are not available. These sources are shown on a color-color diagram in \Cref{fig:ch3_color_color}.

@@

\subsubsection{Group 3 -- Point Source Counterpart With a Bright Nearby Galaxy}
\label{sec:ch3_group3}

@@

\begin{figure}[!t]
\begin{center}
\hspace*{-0.1in}
\scalebox{1.}
{\includegraphics[width=0.9\textwidth]{./figs/chapter3/f4.pdf}}
\caption{Light curves for the 11 sources in our red kilonova contaminant sample, constructed from the ``forced" {\tt DoPhot} PSF photometry (blue circles: $i$-band; orange squares: $z$-band). Nuclear sources are indicated by an (N) in the ID number. The $5\sigma$ limits for non-detections are indicated by triangles.}
\label{fig:ch3_final_lc_red}
\end{center}
\vspace{0.5cm}
\end{figure}

@@

Matching the sources in Group 4 against the external catalogs, we find that all but 13 of these sources have a match in 2MASS, WISE, or Gaia DR1. We show 10 of these 13 unmatched sources on the color-color diagram in \Cref{fig:ch3_color_color} and we find that they are consistent with the stellar locus.\footnote{The remaining 3 sources are not plotted due to a lack of $r$- or $z$-band detections.} Inspecting the magnitude distribution in \Cref{fig:ch3_color_color}, we find that these sources are located at the faint end of our distribution. For example, all of the sources have $r > 20.5$ mag, indicating that they are too faint to be present in the Gaia DR1 catalog. As with Group 3, we conclude that none of these sources can be considered kilonova contaminants, {\it as long as} sufficiently deep template images are available.

@@

@@

\subsection{Color Selection of Kilonova Contaminants}
\label{sec:ch3_kn_search}
We now study the $i-z$ colors of the 48 potential contaminants to identify sources that can mimic either red or blue kilonova emission. The color distribution for all 48 sources is shown in \Cref{fig:ch3_color_dist_final}. The $i-z$ color is computed as a signal-to-noise-weighted average using the ``forced" {\tt DoPhot} photometry from the difference images. We define red kilonova contaminants as those having $i-z \gtrsim 0.5$~mag \citep{BarnesKasen13,CowpBerger15}. For blue kilonova contaminants we require $i-z \lesssim 0.0$~mag, motivated by the models of \cite{Kasen+15} for a BNS merger that results in a HMNS with $t_{\rm HMNS}\gtrsim 100$ ms. These criteria are shown in \Cref{fig:ch3_color_dist_final}. These criteria capture the tails of the color distribution, with over half of the sources (27 of 48) exhibiting $i-z = 0.0-0.5$~mag.

We find 10 sources that have colors expected for blue kilonovae. About half of these sources are located within a pixel of a galaxy nucleus. We show the light curves of all 10 sources in \Cref{fig:ch3_final_lc_blue}. The temporal evolution of blue kilonova models is more rapid, with a shorter duration, than that of red kilonovae. We do not find any systematic trends in light curve behavior for the 10 sources, or when inspecting the nuclear and non-nuclear sources separately.

@@

\subsection{A Search for Contamination on Nightly Timescales}
\label{sec:ch3_kn_short}
We also search our data for sources that could appear as contaminants on the timescales relevant to the short-lived ``neutron precursor," that are speculated to accompany some mergers (\cref{sec:ch3_kn_models}, \citealt{Metzger+15}). We accomplish this by leveraging the rapid cadence of our observations to identify potential contaminants that are detected during a single night, or just a half-night epoch, probing transient and variable events that occur with timescales of $3$ hr to 1 day and $\lesssim3$ hours, respectively. We search for candidates based on their behavior in the ``forced" {\tt DoPhot} photometry as follows:

@@

We find 9 sources with a timescale between 3 hours and 1 day. We perform a manual inspection of this sample, finding 5 genuine sources and 4 that result from image subtraction artifacts. Matching the 5 sources to our template images and external catalogs, we find that all of them have a point-source match in both our template images and Gaia DR1, and hence represent stellar variability or flaring. We find no evidence for extragalactic contamination from sources with a timescale between 3 hours and 1 day.

@@

\begin{figure}[!t]
\begin{center}
\hspace*{-0.1in}
\scalebox{1.}
{\includegraphics[width=0.9\textwidth]{./figs/chapter3/f6.pdf}}
\caption{Plots of recovery efficiency as a function of various fake source injection parameters. The efficiency for the entire population of fake sources is shown as a black line while the efficiency for red kilonova and blue kilonova fake sources are shown as orange and blue lines, respectively. {\it Top Left:} Efficiency as a function of $i$-band magnitude. We note that our efficiency for red kilonova fake sources is higher than our mean efficiency, while the blue kilonova efficiency declines more rapidly compared to the red kilonova fake sources. This is due to the design of our observations, which are aimed at red sources. {\it Top Right:} Efficiency as a function of $i-z$ color. We are more sensitive to objects that are red in $i-z$ color, which drives the efficiency differences seen in the other panels. {\it Bottom Left:} Efficiency as a function of fading ($\delta m$, see text for definition). {\it Bottom Right:} Efficiency as a function of host galaxy separation. We find no dependence on separation.}
\label{fig:ch3_eff_1D}
\end{center}
\end{figure}

>>>>

\section{Introduction}
\label{sec:ch3_intro}
The detection of gravitational wave (GW) events during the first two observing runs of the Advanced Laser Interferometer Gravitational-Wave Observatory (aLIGO) ushered in the era of GW astronomy. The first four announced detections (GW150914, GW151226, GW170104, and GW170814) were due to binary black hole (BBH) mergers, with component masses of $\approx8-36$ M$_\odot$ \citep{LIGOGW150914,LIGOGW151226,LIGOGW170104,LIGOGW170814}.

More recently LIGO announced the first detection of gravitational waves from the inspiral and merger of a binary neutron star \citep[GW170817][]{LIGOGW170817}. This remarkable event was accompanied 1.7s later by a coincident short gamma-ray burst (SGRB) detected with the {\it Fermi} and {\it INTEGRAL} satellites \citep{LIGOGW170817grb,GW170817Fermi,Savchenko+17}. An intense follow-up effort led to the discovery of an associated optical counterpart just 11 hours post-merger \citep{LIGOMMAPaper,Arcavi+17,Coulter+17,GW170817DECam,Valenti+17}.

Initial modeling of this optical emission revealed that it was consistent expectations for a kilonova \citep{Cowp+17,Kilpatrick+17,Tanaka+17,Villar+17b, Tanaka+18}, an isotropic thermal optical/NIR transient powered by the radioactive decay of $r$-process nuclei synthesized in the merger ejecta \citep[see e.g.,][]{Metzger+10,Kasen+13,BarnesKasen13,Metzger2017}. Subsequent studies into the EM counterpart revealed the nature of the radio and X-ray emission \citep{Alexander+17,Davanzo+18,Margutti+17,Margutti+18,Mooley+18,Troja+17} and confirmed the notion of compact object mergers as the progenitors of SGRBs \citep{Fong+13, Berger2014, Fong+15, Fong+17}.

While the case of GW170817 represents a clear success and the dawn of joint GW-EM astronomy, there are many unique aspects of this discovery that should be considered. First, the source was unusually close \citep[$D \approx 40$~Mpc,][]{LIGOGW170817} with a clearly associated host galaxy \citep[NGC 4993,][]{Blanchard+17,Cantiello+18}. Furthermore, the source was unexpectedly bright at discovery \citep[$m_i \approx 17.5$~mag,][]{LIGOMMAPaper,Arcavi+17,Coulter+17,GW170817DECam,Valenti+17} making searches with 1-m class telescopes feasible. It is not clear that future detections by aLIGO will be so fortuitous or if the observed features of the kilonova associated with GW170817 will be ubiquitous. Therefore, in the context of future searches for EM counterparts, wide-field instruments on large telescopes will still be necessary to perform efficient and effect follow-up of GW sources.

For wide-field searches conducted in the optical bands, kilonovae like the one associated with GW170817 are of particular interest due to the isotropic nature of the emission and the expectation that this emission will accompany all BNS mergers, and some of the NS-BH mergers in which the NS is disrupted outside the event horizon \citep{MetzgerBerger12}. The short duration ($\lesssim {\rm week}$) and low luminosity ($L_{\rm bol} \sim {\rm a few} \times 10^{41}$~erg s$^{-1}$ at peak) of the kilonova emission make deep and rapid searches imperative. Coupled with the large search regions, this also requires the development of methodologies to robustly identify kilonovae among the vast numbers of potential contaminating sources. For example, optical follow-up of GW151226, covering hundreds of square degrees, has led to the identification of tens of unrelated transient sources despite being shallower than ideal searches for kilonova emission require \citep[e.g.,][]{GW151226PS1}. Similar wide-field efforts for GW170817 \citep[e.g., DECam][]{GW170817DECam} found 81 transients in a search covering 70 deg$^2$.

@@

\clearpage
The paper is organized as follows. In \cref{sec:ch3_kn_models} we describe the various kilonova models explored in this work. We present the observations and data analysis in \cref{sec:ch3_obs}. In \cref{sec:ch3_search} we discuss the methodologies and results of our search for kilonova contaminants. In \cref{sec:ch3_fakes}, we determine our search efficiency using point sources injected into the images with a wide range of brightness, fading rate, and color. In \cref{sec:ch3_contam_rates} we determine the contamination rate as a function of $i-z$ color and compare our results to those reported from optical follow-up of the current BBH merger events. We discuss the implications of our search in the context of current and future GW detectors and follow-up facilities in \cref{sec:ch3_conc}.

@@

Radioactive decay of the $r$-process elements synthesized during the merger heats the ejecta producing an isotropic, thermal transient \citep{LP98,Rosswog2005, Metzger+10,Tanaka+14}. The combination of low ejecta mass and high ejecta velocity, coupled with the strong optical line blanketing, results in a transient that is faint ($i \approx 23$ and $z \approx 22$ mag at 200 Mpc), red ($i-z \gtrsim 0.5$ mag), and short-lived with a typical duration of $\sim {\rm few}$ days in $z$-band and $\sim{\rm week}$ in $J$-band \citep{BarnesKasen13,Barnes+16,Metzger2017}. In the case of larger opacities \citep[e.g.,][]{Fontes+17}, the transient is expected to peak in the IR (\apx3~$\mu$m), with a duration of \apx10~days \citep{Fontes+17,Wollaeger+17}.

@@

To summarize, kilonova emission with red color $(i-z\gtrsim 0.5)$, a peak brightness of $z\approx22.2$~mag at 200 Mpc, and a duration of $\sim \rm few$ days is expected to be ubiquitous. Blue kilonova emission due to a surviving HMNS is expected to be bluer ($i-z \lesssim 0$~mag), similarly faint ($r \approx 22$~mag) and shorter in duration ($\lesssim 1-2$ days). This picture was confirmed by modeling of the optical emission associated with GW170817. Successfully fitting the data required models comprised of at least two components (e.g., `red' and `blue' emission) with their own distinct ejecta properties \citep{Cowp+17,Kilpatrick+17,Tanaka+17,Villar+17b, Tanaka+18}. The observed behavior and composition of the dynamical ejecta and associated red kilonova emission is robust and ubiquitous, however the nature and observability of the blue kilonova emission depends both on the fraction of cases in which a HMNS survives (unknown) and on geometrical effects ($\lesssim 50\%$).  Finally, emission due to free neutrons will be similar in brightness and color to the blue kilonova emission, but with a timescale of only $\sim \rm few$ hours; the prevalence of this signal is uncertain.  The observations described in this paper address contamination in all of these cases.

@@

\singlespace
\begin{deluxetable}{lcc}
\tabletypesize{\footnotesize}
\tablecolumns{3}
\tablewidth{0pt}
\captionsetup{justification=centering}
\tablecaption{Field Pointing Centers
\label{tab:ch3_pointings}}
\tablehead{
    \colhead{Pointing} &
    \colhead{R.A.} &
    \colhead{Decl.}}
\startdata
Antlia 1 & 10:19:58.5  & $-$30:55:24 \\
Antlia 2 & 10:19:44.5  & $-$33:07:24 \\
Antlia 3 & 10:19:30.5  & $-$35:19:24 \\
Antlia 4 & 10:19:10.5  & $-$37:31:24 \\
Antlia 5 & 10:30:03.5  & $-$30:55:24 \\
Antlia 6 & 10:30:03.5  & $-$33:07:24 \\
Antlia 7 & 10:30:03.5  & $-$35:19:24 \\
Antlia 8 & 10:30:03.5  & $-$37:31:24 \\
Antlia 9 & 10:40:08.5  & $-$30:55:24 \\
Antlia 10 & 10:40:22.5  & $-$33:07:24 \\
Antlia 11 & 10:40:38.5  & $-$35:19:24 \\
Antlia 12 & 10:40:57.5  & $-$37:31:24 \\
Antlia 13 & 10:50:12.5  & $-$30:55:24 \\
Antlia 14 & 10:50:41.5  & $-$33:07:24 \\
Antlia 15 & 10:51:13.5  & $-$35:19:24 \\
Antlia 16 & 10:51:50.5  & $-$37:31:24 \\
Antlia 17 & 11:00:17.5  & $-$30:55:24 \\
Antlia 18 & 11:01:00.5  & $-$33:07:24 \\
Antlia 19 & 11:01:48.5  & $-$35:19:24 \\
Antlia 20 & 11:02:44.5  & $-$37:31:24 \\
Antlia 21 & 10:36:36.0  & $-$27:31:04 \\
\enddata
\tablecomments{Central J2000 coordinates for all 21 fields used in this analysis.}
\end{deluxetable}
\doublespace

\clearpage
We processed the data using {\tt photpipe}, an image processing pipeline used by several previous time domain surveys (see e.g., \citealt{Rest+05,Rest+14}) to perform single-epoch image processing, image subtraction, and candidate identification. The single epoch processing began with the raw images and appropriate calibration files obtained from the NOAO archive\footnote{\url{http://archive.noao.edu/}} and initial image reduction (e.g., bias subtraction, cross-talk corrections, flat-fielding). We performed astrometric calibration relative to the 2MASS $J$-band point-source catalog. Search and template images were deprojected into a tangential plane and pairs of $i$- and $z$- band images from a single epoch were coadded using {\tt SWARP} \citep{Bertin+02}.

@@

\clearpage
In addition to the photometry performed by {\tt photpipe}, we also constructed secondary catalogs for all sources identified in the $griz$-band template epochs. This was accomplished using the {\tt Source Extractor} ({\tt SExtractor}) photometry package in single-image mode \citep{BertinArnouts96}. We also performed forced aperture photometry in the template epoch at the position of each candidate identified by {\tt photpipe}. This approach helps to identify the presence of flux in the template images for objects not detected by {\tt SExtractor} which is indicative of image artifacts and defects. We use these additional template catalogs as a useful tool for candidate classification and artifact rejection (see \cref{sec:ch3_source_cuts}).

@@

\clearpage
\section{A Search for Optical Transients}
\label{sec:ch3_search}

@@

\clearpage
We note that at the time these data were taken CCD \#61 (N30) was not functioning. Additionally, during processing of the candidate list we observed severe data quality issues with CCDs \#16 (S17) and \#44 (N13) that produced a number of image artifacts and erroneous detections several orders of magnitude larger than in the other CCDs. Consequently, these three CCDs have been excluded from the analysis. This results in a $\approx 5\%$ loss of sky coverage as discussed in \cref{tab:ch3_search}.

@@

In Groups 3 and 4, the choice of galaxy brightness and matching radius are motivated by observations of the host galaxies of Short GRBs, which have luminosities of $\gtrsim 0.1$ L$^\star$ \citep{FongBerger13,Fong+13,Berger2014}. At the ALV design-sensitivity detection range $(\sim 200$ Mpc) such galaxies will be brighter than 17.5 mag.

\clearpage
We then select the matching radius such that the probability of the point source and galaxy being associated by chance, the ``probability of chance coincidence", is $P_{\rm cc}\lesssim 0.1$. This is given by:

@@

@@

We note that the single source in \Cref{fig:ch3_offset_hist} with an offset of $\approx6.6\arcsec$, was originally identified with an offset of $0.14\arcsec$. However, manual inspection revealed that the transient had not faded away in the template causing {\tt SExtractor} deblending to detect a low stellarity point source still present on top of the galaxy light distribution. We correct this offset manually, but leave this source in Group 2 as that is the {\it original} classification.

\clearpage
\subsubsection{Groups 3 and 4 -- Point Source Counterparts}
\label{sec:ch3_pointsources}
We are also interested in those candidates that have a coincident point source counterpart in the template images. The presence of such a counterpart disqualifies the source as a kilonova contaminant, but this determination relies on the existence of late-time (or pre-existing) templates.  Therefore, it is still meaningful to construct a census of these sources to gain a better understanding of potential contamination in real-time GW follow-up observations, especially if pre-existing template images are not available. These sources are shown on a color-color diagram in \Cref{fig:ch3_color_color}.

\begin{figure}[!t]
\begin{center}
\hspace*{-0.1in}
\scalebox{1.}
{\includegraphics[width=\textwidth]{./figs/chapter3/f2alt.pdf}}
\caption{{\it Left:} $r-i$ vs. $i-z$ color-color diagram for the template counterparts identified in Groups 3 (orange stars; \cref{sec:ch3_group3}) and 4 (blue circles; \cref{sec:ch3_group4}). The black contours indicate the stellar locus in our images. We find that the majority of sources are consistent with the stellar locus indicating a variable star origin.
{\it Middle:} Same as top, but only showing those 14 sources that do not have a match in the Gaia DR1 catalog. The majority of sources are still consistent with the stellar locus, indicating that they are stellar but simply too faint to appear in the Gaia catalog.
{\it Bottom:} Distribution of $r$-band magnitudes for sources in Groups 3 and 4. The solid lines indicate sources that have a counterpart in the Gaia DR1 catalog, while dashed lines indicate sources that do not have a catalog match.}
\label{fig:ch3_color_color}
\end{center}
\end{figure}

@@

\clearpage
\subsubsection{Group 3 -- Point Source Counterpart With a Bright Nearby Galaxy}
\label{sec:ch3_group3}

@@

@@

Matching the sources in Group 4 against the external catalogs, we find that all but 13 of these sources have a match in 2MASS, WISE, or Gaia DR1. We show 10 of these 13 unmatched sources on the color-color diagram in \Cref{fig:ch3_color_color} showing that they are consistent with the stellar locus.\footnote{The remaining 3 sources are not plotted due to a lack of $r$- or $z$-band detections.} Inspecting the magnitude distribution in \Cref{fig:ch3_color_color}, we find that these sources are located at the faint end of our distribution. All of the sources exhibit $r > 20.5$ mag, indicating that they are too faint to be present in the Gaia DR1 catalog. As with Group 3, we conclude that none of these sources can be considered kilonova contaminants, {\it as long as} sufficiently deep template images are available.

@@

\clearpage
\subsection{Color Selection of Kilonova Contaminants}
\label{sec:ch3_kn_search}
We now study the $i-z$ colors of the 48 potential contaminants to identify sources that can mimic either red or blue kilonova emission. The color distribution for all 48 sources is shown in \Cref{fig:ch3_color_dist_final}. The $i-z$ color is computed as a signal-to-noise-weighted average using the ``forced" {\tt DoPhot} photometry from the difference images. We define red kilonova contaminants as those having $i-z \gtrsim 0.5$~mag \citep{BarnesKasen13,CowpBerger15}. For blue kilonova contaminants we require $i-z \lesssim 0.0$~mag, motivated by the models of \cite{Kasen+15} for a BNS merger that results in a HMNS with $t_{\rm HMNS}\gtrsim 100$ ms. These criteria are shown in \Cref{fig:ch3_color_dist_final}. These criteria capture the tails of the color distribution, with over half of the sources (27 of 48) exhibiting $i-z = 0.0-0.5$~mag.

\begin{figure}[!t]
\begin{center}
\hspace*{-0.1in}
\scalebox{1.}
{\includegraphics[width=\textwidth]{./figs/chapter3/f3.pdf}}
\caption{{\it Left:} $i-z$ color distribution for the 48 sources considered in this analysis. The 11 sources identified as red kilonova contaminants are plotted in orange (\cref{sec:ch3_kn_search}), while the 10 sources identified as blue kilonova contaminants are shown in blue (\cref{sec:ch3_kn_search}). The remaining 27 sources from Group 2 are plotted in grey dashed lines.
{\it Right:} The number of sources recovered by performing an $i-z$ color cut on the sample of 48 sources in Group 2. The orange line gives the number of sources redder than a given $i-z$ color. The blue line indicates the number of sources bluer than a given color cut. The vertical black lines indicate our nominal cuts of $i-z < 0$~mag and $i-z > 0.5$~mag, for blue and red kilonovae, respectively. The sharp increase in the number of sources as the chosen color threshold is relaxed can be clearly seen.}
\label{fig:ch3_color_dist_final}
\end{center}
\end{figure}

\begin{figure}[!t]
\begin{center}
\hspace*{-0.1in}
\scalebox{1.}
{\includegraphics[width=0.825\textwidth]{./figs/chapter3/f4.pdf}}
\caption{Light curves for the 11 sources in our red kilonova contaminant sample, constructed from the ``forced" {\tt DoPhot} PSF photometry (blue circles: $i$-band; orange squares: $z$-band). Nuclear sources are indicated by an (N) in the ID number. The $5\sigma$ limits for non-detections are indicated by triangles.}
\label{fig:ch3_final_lc_red}
\end{center}
\vspace{0.5cm}
\end{figure}

@@

\clearpage
We find 10 sources that have colors expected for blue kilonovae. About half of these sources are located within a pixel of a galaxy nucleus. We show the light curves of all 10 sources in \Cref{fig:ch3_final_lc_blue}. The temporal evolution of blue kilonova models is more rapid, with a shorter duration, than that of red kilonovae. We do not find any systematic trends in light curve behavior for the 10 sources, or when inspecting the nuclear and non-nuclear sources separately.

@@

\clearpage
\subsection{A Search for Contamination on Nightly Timescales}
\label{sec:ch3_kn_short}
We also search our data for sources that could appear as contaminants on the timescales relevant to the short-lived ``neutron precursor," that are speculated to accompany some mergers (\cref{sec:ch3_kn_models}, \citealt{Metzger+15}). We accomplish this by leveraging the rapid cadence of our observations to identify potential contaminants that are detected during a single night, or just a half-night epoch, probing transient and variable events that occur with timescales of $3$ hr to 1 day and $\lesssim3$ hours, respectively. We search for candidates based on their behavior in the ``forced" {\tt DoPhot} photometry as follows:

@@

\clearpage
We find 9 sources with a timescale between 3 hours and 1 day. We perform a manual inspection of this sample, finding 5 genuine sources and 4 that result from image subtraction artifacts. Matching the 5 sources to our template images and external catalogs, we find that all of them have a point-source match in both our template images and Gaia DR1, and hence represent stellar variability or flaring. We find no evidence for extragalactic contamination from sources with a timescale between 3 hours and 1 day.

@@
